### 2025-03-14 paper/sections/experiments.qmd
Shorten to slightly less than 16 pages by restructuring

In this section, we present experiments that we have conducted in order to answer the following research questions:

1. To what extent does our proposed counterfactual training objective in @eq-obj induce models to learn plausible explanations?
2. To what extent does our proposed counterfactual training objective in @eq-obj yield more favorable algorithmic recourse outcomes in the presence of actionability constraints?
3. What are the effects of hyperparameter selection with respect to @eq-obj?

@@

### Evaluation

@@

### Plausibility {#sec-plaus}

@tbl-main presents our main empirical findings. The top five rows show the percentage reduction in implausibility according to @eq-impl-dist for varying degrees of the energy penalty used for *ECCo* at test time. The following row shows the reduction in implausibility as measured by @eq-impl-div and aggregated across all test specifications of *ECCo*. The following two rows show the test accuracies for the model trained with CT and conventionally trained models ("vanilla").

We observe that for all datasets except *OL* and across all test settings, the average distance of counterfactuals from observed samples in the target class is reduced, indicating improved plausibility. The magnitude of improvements varies by dataset: for the simple synthetic datasets, distance reductions range from around 20-40% (*LS*, *Moon*) to almost 60% (*Circ*). For the real-world tabular datasets, improvements are generally smaller but still substantial in many cases with around 10-15% for *CH*, 11-28% for *GMSC*, 7-8% for *Cred* and around 3% for *Adult*. For our only vision dataset (*MNIST*), distances are reduced by up to 9%. The results for our proposed divergence metric are qualitatively similar, but generally even more pronounced: for the *Circ* dataset, implausibility is reduced by almost 94% to virtually zero as we verified by looking at the absolute outcome. Improvements for other datasets range from 28% (*Moon*) to 78% (*GMSC*). For *OL* the reduction is negative, consistent with the distance-based metric. *MNIST* is the only dataset for which the two metrics disagree.

@@

### Actionability {#sec-act}

@@

### Impact of hyperparameter settings {#sec-hyperparameters}
We test in-depth the impact of three types of hyperparameters; our complete results are in the appendix.

***Hyperparameters of the CE generators.*** First, we observe that CT is highly sensitive to hyperparameter settings but (a) there are manageable patterns and (b) we can typically identify settings that improve either plausibility or cost, and commonly both of them at the same time. Second, we note that the choice of a CE generator has a major impact on the results. For example, *REVISE* tends to perform the worst, most likely because it uses a surrogate VAE to generate counterfactuals which impedes faithfulness [@altmeyer2024faithful]. Third, increasing $T$, the maximum number of steps, generally yields better outcomes because more CEs can mature in each training epoch. Fourth, the impact of $\tau$, the required decision threshold is more difficult to predict. On "harder" datasets it may be difficult to satisfy high $\tau$ for any given sample (i.e., also factuals) and so increasing this threshold does not seem to correlate with better outcomes. In fact, we have generally found that a choice of $\tau=0.5$ leads to optimal results because it is associated with high proportions of mature counterfactuals.

***Hyperparameters for penalties.*** We find that the strength of the energy regularization, $\lambda_{\text{reg}}$ is highly impactful; energy must be sufficiently regularized to avoid poor performance in terms of decreased plausibility and increased costs. The sensitivity with respect to $\lambda_{\text{div}}$ and $\lambda_{\text{adv}}$ is much less evident. While high values of $\lambda_{\text{reg}}$ may increase the variability in outcomes when combined with high values of $\lambda_{\text{div}}$ or $\lambda_{\text{adv}}$, this effect is not very pronounced.

***Other hyperparameters.*** We observe that the effectiveness and stability of CT is positively associated with the number of counterfactuals generated during each training epoch. We also confirm that a higher number of training epochs is beneficial. Interestingly, we observed desired improvements in explainability when CT was combined with conventional training and applied only for the final 50% of epochs of the complete training process. Put differently, CT may be a way to improve the explainability of models in a fine-tuning manner.

>>>>

Here we present the experiments conducted to answer our research questions:

1. To what extent does our counterfactual training objective as defined in @eq-obj induce models to learn plausible explanations?
2. To what extent does the CT objective produce more favorable algorithmic recourse outcomes in the presence of actionability constraints?
3. What are the effects of hyperparameter selection wrt. the CT objective?

@@

@@

### Plausibility. {#sec-plaus}

@tbl-main presents our main empirical findings. The top five rows show the percentage reduction in implausibility according to @eq-impl-dist for varying degrees of the energy penalty used for *ECCo* at test time. Row 6 shows the reduction in implausibility as measured by @eq-impl-div and aggregated across all test specifications of *ECCo*. Rows 7 and 8 show the test accuracies for the model trained with CT and conventionally trained models ("vanilla").

For all datasets except *OL* and across all test settings, the average distance of CEs from observed samples in the target class is reduced, indicating improved plausibility. The magnitude of improvements varies. For the simple synthetic datasets, distance reductions range from around 20-40% (*LS*, *Moon*) to almost 60% (*Circ*). For the real-world tabular datasets, improvements tend to be smaller but still substantial, with around 10-15% for *CH*, 11-28% for *GMSC*, 7-8% for *Cred* and around 3% for *Adult*. For our only vision dataset (*MNIST*), distances are reduced by up to 9%. The results for our proposed divergence metric are qualitatively similar, but generally even more pronounced: for the *Circ* dataset, implausibility is reduced by almost 94% to virtually zero as we verified by the absolute outcome. Improvements for other datasets range from 28% (*Moon*) to 78% (*GMSC*). For *OL* the reduction is negative, consistent with the distance-based metric. *MNIST* is the only dataset for which the two metrics disagree.

@@

### Actionability. {#sec-act}

@@

### Impact of hyperparameter settings. {#sec-hyperparameters}
We test the impact of three types of hyperparameters; our complete results are in the supplementary appendix.  
***Hyperparameters of the CE generators.*** First, we observe that CT is highly sensitive to hyperparameter settings but (a) there are manageable patterns and (b) we can typically identify settings that improve either plausibility or cost, and commonly both of them at the same time. Second, we note that the choice of a CE generator has a major impact on the results. For example, *REVISE* tends to perform the worst, most likely because it uses a surrogate VAE to generate counterfactuals which impedes faithfulness [@altmeyer2024faithful]. Third, increasing $T$, the maximum number of steps, generally yields better outcomes because more CEs can mature in each training epoch. Fourth, the impact of $\tau$, the required decision threshold is more difficult to predict. On "harder" datasets it may be difficult to satisfy high $\tau$ for any given sample (i.e., also factuals) and so increasing this threshold does not seem to correlate with better outcomes. In fact, we have generally found that a choice of $\tau=0.5$ leads to optimal results because it is associated with high proportions of mature counterfactuals.  
***Hyperparameters for penalties.*** We find that the strength of the energy regularization, $\lambda_{\text{reg}}$ is highly impactful; energy must be sufficiently regularized to avoid poor performance in terms of decreased plausibility and increased costs. The sensitivity with respect to $\lambda_{\text{div}}$ and $\lambda_{\text{adv}}$ is much less evident. While high values of $\lambda_{\text{reg}}$ may increase the variability in outcomes when combined with high values of $\lambda_{\text{div}}$ or $\lambda_{\text{adv}}$, this effect is not very pronounced.  
***Other hyperparameters.*** We observe that the effectiveness and stability of CT is positively associated with the number of counterfactuals generated during each training epoch. We also confirm that a higher number of training epochs is beneficial. Interestingly, we observed desired improvements in explainability when CT was combined with conventional training and applied only for the final 50% of epochs of the complete training process. Put differently, CT may be a way to improve the explainability of models in a fine-tuning manner.

### 2025-03-14 paper/sections/intro.qmd
Shorten to slightly less than 16 pages by restructuring

**Specifically, we propose counterfactual training (CT)**: a novel training regime that aligns learned representations with plausible explanations compliant with user requirements. The remainder of this paper is structured as follows. @sec-lit presents related work, focusing on the link between adversarial examples and counterfactual explanations. Then follow our main contributions:

1. In @sec-method, we introduce our methodological framework and show theoretically that it can be used to enforce global actionability constraints.  
2. In @sec-experiments, through extensive experiments we demonstrate that CT substantially improves explainability without sacrificing predictive performance.

We discuss the challenges in @sec-discussion and conclude in @sec-conclusion that CT is a promising approach towards making opaque models more trustworthy.

>>>>

**Specifically, we propose counterfactual training (CT)**: a novel training regime that aligns learned representations with plausible explanations respecting actionability constraints. The remainder of this paper is structured as follows. @sec-lit presents related work, focusing on the link between adversarial examples and counterfactual explanations. Then follow our two principal contributions. In @sec-method, we introduce our methodological framework and show theoretically that it can be employed to enforce global actionability constraints. In @sec-experiments, through extensive experiments, we demonstrate that CT substantially improves explainability without sacrificing predictive performance. We discuss the challenges in @sec-discussion and conclude in @sec-conclusion that CT is a promising approach towards making opaque models more trustworthy.

### 2025-03-14 paper/sections/method.qmd
Shorten to slightly less than 16 pages by restructuring

@grathwohl2020your observe that any classifier can be re-interpreted as a joint energy-based model (JEM) that learns to discriminate output classes conditional on the observed (training) samples from $p(\mathbf{x})$ and the generated samples from $p_\theta(\mathbf{x})$. The authors show that JEMs can be trained to perform well at both tasks by directly maximizing the joint log-likelihood factorized as $\log p_\theta(\mathbf{x},\mathbf{y})=\log p_\theta(\mathbf{y}|\mathbf{x}) + \log p_\theta(\mathbf{x})$. The first term can be optimized using conventional cross-entropy as in @eq-obj. Then, to optimize $\log p_\theta(\mathbf{x})$ @grathwohl2020your minimize the contrastive divergence between these observed samples from $p(\mathbf{x})$ and generated samples from $p_\theta(\mathbf{x})$.

@@

From a Bayesian perspective, setting $\mathbf{x}^+[d] := \mathbf{x}^\prime[d]$ can be understood as assuming a point mass prior for $p(\mathbf{x}^+)$ with respect to feature $d$. Intuitively, we think of this simply in terms ignoring implausibility costs with respect to immutable features, which effectively forces the model to instead seek plausibility with respect to the remaining features. This in turn results in lower overall sensitivity to immutable features, which we demonstrate empirically for different classifiers in @sec-experiments. Under certain conditions, this results holds theoretically:^[For the proof, see the supplementary appendix.]

@@

It is worth highlighting that \mbox{Prp. \ref{prp-mtblty}} assumes independence of features. This raises a valid concern about the effect of protecting immutable features in the presence of proxies that remain unprotected. We address this in @sec-discussion.

@@

We have simulated this scenario using synthetic data with two independent features and Gaussian class-conditional densities in @fig-poc. The four panels in @fig-poc show the outcomes for different training procedures using the same model architecture each time (a linear classifier). In each case, we show the decision boundary (in green) and the training data colored according to their ground-truth label: orange points belong to the target class, $y^+=2$, blue points belong to the non-target class, $y^-=1$. Stars indicate counterfactuals in the target class generated at test time using generic gradient descent until convergence.

In panel (a), we have trained our model conventionally, and we do not impose mutability constraints at test time. The generated counterfactuals are all valid, but not plausible: they do not comply with the distribution of the factual samples in the target class to the point where they are clearly distinguishable from the ground-truth data. In panel (b), we have trained our model with CT, once again without any mutability constraints. We observe that the counterfactuals are highly plausible, meeting the first objective of Def. \ref{def-explainability}.

In panel (c), we have used conventional training again, this time imposing the mutability constraint on *age* at test time. Counterfactuals are valid but involve some substantial reductions in *debt* for some individuals (very young applicants). By comparison, counterfactual paths are shorter on average in panel (d), where we have used CT and protected the immutable feature as described in @sec-constraints. We observe that due to the classifier's lower sensitivity to *age*, recourse recommendations with respect to *debt* are much more homogenous and do not disproportionately punish younger individuals. The counterfactuals are also plausible with respect to the mutable feature. Thus, we consider the model in panel (d) as the most explainable according to Def. \ref{def-explainability}.

>>>>

As observed by @grathwohl2020your, any classifier can be re-interpreted as a joint energy-based model (JEM) that learns to discriminate output classes conditional on the observed (training) samples from $p(\mathbf{x})$ and the generated samples from $p_\theta(\mathbf{x})$. The authors show that JEMs can be trained to perform well at both tasks by directly maximizing the joint log-likelihood factorized as $\log p_\theta(\mathbf{x},\mathbf{y})=\log p_\theta(\mathbf{y}|\mathbf{x}) + \log p_\theta(\mathbf{x})$. The first term can be optimized using conventional cross-entropy as in @eq-obj. Then, to optimize $\log p_\theta(\mathbf{x})$ @grathwohl2020your minimize the contrastive divergence between these observed samples from $p(\mathbf{x})$ and generated samples from $p_\theta(\mathbf{x})$.

@@

From a Bayesian perspective, setting $\mathbf{x}^+[d] := \mathbf{x}^\prime[d]$ can be understood as assuming a point mass prior for $p(\mathbf{x}^+)$ with respect to feature $d$. Intuitively, we think of this simply in terms ignoring implausibility costs with respect to immutable features, which effectively forces the model to instead seek plausibility with respect to the remaining features. This in turn results in lower overall sensitivity to immutable features, which we demonstrate empirically for different classifiers in @sec-experiments. Under certain conditions, this result holds theoretically:^[For the proof, see the supplementary appendix.]

@@

It is worth highlighting that \mbox{Proposition \ref{prp-mtblty}} assumes independence of features. This raises a valid concern about the effect of protecting immutable features in the presence of proxies that remain unprotected. We address this in @sec-discussion.

@@

We have simulated this scenario using synthetic data with two independent features and Gaussian class-conditional densities in @fig-poc. The four panels show the outcomes for different training procedures using the same model architectures (a linear classifier). In panels (a) and (c) we have trained the models conventionally, while in panels (b) and (d) we used CT. Only in panels (c) and (d) do we impose the mutability constraint on *age* at test time. In each case, we show the decision boundary (in green) and the training data colored according to their ground-truth label: orange points belong to the target class, $y^+=2$, blue points belong to the non-target class, $y^-=1$. Stars indicate CEs in the target class generated at test time using generic gradient descent until convergence.

In all cases the counterfactuals are valid but their quality differs. In panel (a), they are not plausible: they do not comply with the distribution of the factuals in $y^+$ to the point where they form a clearly distinguishable cluster. In panel (b), they are highly plausible, meeting the first objective of Def. \ref{def-explainability}. In panel (c), the CEs involve substantial reductions in *debt* for younger applicants. By comparison, counterfactual paths are shorter on average in panel (d) where we have protected the immutable *age* as described in @sec-constraints. Due to the classifier's lower sensitivity to *age*, recommendations with respect to *debt* are much more homogenous and do not unfairly punish younger individuals. These counterfactuals are also plausible with respect to the mutable feature. Thus, we consider the model in panel (d) as the most explainable according to Def. \ref{def-explainability}.

### 2025-03-14 paper/sections/related.qmd
Shorten to slightly less than 16 pages by restructuring

In the natural language processing (NLP) domain, counterfactuals have similarly been used to improve models through data augmentation. @wu2021polyjuice propose *Polyjuice*, a general-purpose counterfactual generator for language models. They demonstrate empirically that the augmentation of training data through *Polyjuice* counterfactuals improves robustness in a number of NLP tasks. @balashankar2023improving similarly use *Polyjuice* to augment NLP datasets through diverse counterfactuals and show that classifier robustness improves by up to 20%. Finally, @luu2023counterfactual introduce Counterfactual Adversarial Training (CAT), which also aims at improving generalization and robustness of language models through a three-step procedure. First, the authors identify training samples that are subject to high predictive uncertainty. Second, they generate counterfactual explanations for those samples. Finally, they fine-tune the given language model on the augmented dataset that includes the generated counterfactuals.

@@

Two recent works are closely related to ours in that they use counterfactuals during training with the explicit goal of affecting certain properties of the post-hoc counterfactual explanations. Firstly, @ross2021learning propose a way to train models that guarantee individual recourse to some positive target class with high probability. Their approach builds on adversarial training by explicitly inducing susceptibility to targeted adversarial examples for the positive class. Additionally, the proposed method allows for imposing a set of actionability constraints ex-ante. For example, users can specify that certain features (e.g., *age*, *gender*) are immutable. Secondly, @guo2023counternet are the first to propose an end-to-end training pipeline that includes counterfactual explanations as part of the training procedure. In particular, they propose a specific network architecture that includes a predictor and CE generator network, where the parameters of the CE generator network are learnable. Counterfactuals are generated during each training iteration and fed back to the predictor network. In contrast to @guo2023counternet, we impose no restrictions on the neural network architecture at all.

>>>>

In the natural language processing (NLP) domain, CEs have also been used to improve models through data augmentation. @wu2021polyjuice propose *Polyjuice*, a general-purpose counterfactual generator for language models. The authors empirically demonstrate that the augmentation of training data with their method improves robustness in a number of NLP tasks. @balashankar2023improving similarly use *Polyjuice* to augment NLP datasets through diverse counterfactuals and show that classifier robustness improves by up to 20%. Finally, @luu2023counterfactual introduce Counterfactual Adversarial Training (CAT), which also aims at improving generalization and robustness of language models through a three-step procedure: the authors identify training samples that are subject to high predictive uncertainty, generate CEs for those samples, and then fine-tune the language model on a dataset augmented with the CEs.

@@

Two recent works are closely related to ours in that they use counterfactuals during training with the explicit goal of affecting certain properties of the post-hoc counterfactual explanations. Firstly, @ross2021learning propose a way to train models that guarantee individual recourse to some positive target class with high probability. Their approach builds on adversarial training by explicitly inducing susceptibility to targeted adversarial examples for the positive class. Additionally, the proposed method allows for imposing a set of actionability constraints ex-ante. For example, users can specify that certain features are immutable. Secondly, @guo2023counternet are the first to propose an end-to-end training pipeline that includes CEs as part of the training procedure. In particular, they propose a specific network architecture that includes a predictor and CE generator network, where the parameters of the CE generator network are learnable. Counterfactuals are generated during each training iteration and fed back to the predictor network. In contrast to @guo2023counternet, we impose no restrictions on the ANN architecture at all.
